\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models}

### 2021-11-15 Manuscripts/Manuscript_by_chapters-SIM_Revisions/Full_document_SIM_No_Appendix.tex
updated GAM section

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data. The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data.\textsuperscript{\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}} Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} However, LMEMs impose restrictions in the distribution of the errors of the random effects, which need to be normally distributed and independent.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}} And even more importantly, LMEMs also assume by default a linear relationship between the response and time,\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} making them unsuitable to analyze data non-linear with non-linear trends (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section \{\#GAM-theory\}) .

@@

In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively. Independent errors \(\varepsilon_{ijt}\) represent random variation not explained by the \emph{fixed} effects, and are assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug,'' or ``saline'' vs.~``chemotherapy''). Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B). With this notation, the linear model then can be expressed as

@@

\hypertarget{the-linear-mixed-model-lmem-case}{%
\subsubsection{The Linear Mixed Model (LMEM) Case}\label{the-linear-mixed-model-lmem-case}}

@@

Comparing the fitted responses of the LMEM and the rm-ANOVA models used in the simulated quadratic data (Figure \ref{fig:l-q-response}E, F) indicates that the models are not capturing the changes within each group. Specifically, note that the fitted mean response of both models shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured. The LMEM is only able to account for between-subject variation by providing estimates for each subject (Figure \ref{fig:l-q-response}F), but both models are unable to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare Figure \ref{fig:l-q-response}D with Figure \ref{fig:l-q-response}E, F). If such a change has important physiological implications, both rm-ANOVA and LMEMs omit it from the fitted mean response. Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.

This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data. The models fitted to the simulated data were an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response. In the following section, we present generalized additive models (GAMs) as a data-driven alternative method to analyze longitudinal non-linear data that overcomes the linearity assumption.

\hypertarget{GAM-theory}{%
\section{GAMs as a special case of Generalized Linear Models}\label{GAM-theory}}

\hypertarget{gams-and-basis-functions}{%
\subsection{GAMs and Basis Functions}\label{gams-and-basis-functions}}

Generalized linear models (GLMs) are a family of models (which include rm-ANOVA and LMEMs) that fit a linear response function to data that may not have normally distributed errors.\textsuperscript{\protect\hyperlink{ref-nelder1972}{51}} In contrast, GAMs are a family of regression-based methods for estimating smoothly varying trends and are a broader class of models that contain the GLM family as a special case\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{52}}. A GAM model can be written as:

@@

Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, the change of \(y_{ijt}\) over time is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual error.

@@

The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions. A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(x_t\mid \beta_j)\) is thought to exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\). The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{53}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions. In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.

Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{54}} Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.

Cubic splines (CS), are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness. The use of CS as smoothers in GAMs was discussed within the original GAM framework,\textsuperscript{\protect\hyperlink{ref-hastie1987}{52}} but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit.\textsuperscript{\protect\hyperlink{ref-wood2003}{55}} A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate.\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{55}}

In contrast, thin regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner.\textsuperscript{\protect\hyperlink{ref-wood2003}{55}} Like TPS, TPRS only require the number of basis to be used to create the smoother (for mathematical details on both TPS and TPRS see refs.\textsuperscript{\protect\hyperlink{ref-wood2003}{55}} and).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}

@@

\hypertarget{a-bayesian-interpretation-of-gams}{%
\section{A Bayesian interpretation of GAMs}\label{a-bayesian-interpretation-of-gams}}

Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{56}} In the case of data that shows non-linear trends, the belief that the \emph{true} trend of the data is likely to be smooth rather than extremely ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language).\textsuperscript{\protect\hyperlink{ref-miller2019}{57}} Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (point-wise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}

@@

Contrary to this, an \emph{across the function} CI (like those estimated for GAMs using \emph{mgcv}) contains the true function through the entire range of the covariates (which would be time in the case of longitudinal data). For GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, the percentage of CIs entirely containing the true function over the entire time period would be close to the nominal value (i.e., if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function), thereby allowing more robust estimates from the model. In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{57}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{58}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.

@@

An interaction effect is typically the main interest in longitudinal biomedical data, as it takes into account treatment, time, and their combination. In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time. Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry. A detailed description on basis functions and correlations can be found in.\textsuperscript{\protect\hyperlink{ref-hefley2017}{53}}

@@

When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B). Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{59}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{60}}

@@

This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model. We acknowledge that polynomial effects can be used in LMEMs, and that it could be argued that in Section \ref{simulation} a LMEM with quadratic effects could have been fitted, but in reality the true function in the data is not known; using polynomial effects causes more questions than answers in this case (i.e., what is the appropriate polynomial effect to use? quadratic?, cubic?, what is the biological interpretation of a polynomial effect?). Moreover, polynomial effects have other associated issues that we have described in Section \{\#GAM-theory\}.

Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{61}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) value, which is not meaningful in the case of a Type III error.

In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{53}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{62}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model (i.e., polynomial effects in LMEMs).

@@

We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible. In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{63},\protect\hyperlink{ref-weissgerber2018}{64}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field. Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research. We have made all the data and code used in this paper accessible, and we hope that this will encourage other researchers to do the same with future projects.

@@

\leavevmode\vadjust pre{\hypertarget{ref-nelder1972}{}}%
\CSLLeftMargin{51. }
\CSLRightInline{Nelder JA, Wedderburn RWM. Generalized linear models. \emph{Journal of the Royal Statistical Society Series A (General)}. 1972;135(3):370. doi:\href{https://doi.org/10.2307/2344614}{10.2307/2344614}}

\leavevmode\vadjust pre{\hypertarget{ref-hastie1987}{}}%
\CSLLeftMargin{52. }
\CSLRightInline{Hastie T, Tibshirani R. Generalized additive models: Some applications. \emph{Journal of the American Statistical Association}. 1987;82(398):371-386. doi:\href{https://doi.org/10.1080/01621459.1987.10478440}{10.1080/01621459.1987.10478440}}

>>>>

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data. The recognition on the shortcomings of rm-ANOVA is exemplified by the use of linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data.\textsuperscript{\protect\hyperlink{ref-skala2010}{8},\protect\hyperlink{ref-vishwanath2009}{16}} Briefly, LMEMs incorporate \emph{fixed effects}, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and \emph{random effects}, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject.\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} However, LMEMs impose restrictions in the distribution of the random effects, which need to be independent.\textsuperscript{\protect\hyperlink{ref-gueorguieva2004}{13},\protect\hyperlink{ref-barr2013}{31}} And even more importantly, LMEMs also assume by default a linear relationship between the response and time\textsuperscript{\protect\hyperlink{ref-pinheiro2006}{15}} (polynomial effects can be used with LMEMs, but this approach has its own shortcomings as we discuss in Section \ref{GAM-theory}) .

@@

In this model \(y_{ijt}\) is the response for subject \(i\), in treatment group \(j\) at time \(t\), which can be decomposed in a mean value \(\beta_0\), \emph{fixed effects} of time (\(time_t\)), treatment (\(treatment_j\)) and their interaction \(time_t*treatment_j\) which have linear slopes given by \(\beta_1, \beta_2\) and \(\beta_3\), respectively. Independent errors \(\varepsilon_{ijt}\) represent \textbf{random variation from the sampling process } assumed to be \(\sim N(0,\sigma^2)\) (independently and identically normally distributed with mean zero and variance \(\sigma^2\)).
In a biomedical research context, suppose two treatments groups are used in a study (e.g., ``placebo'' vs.~``novel drug,'' or ``saline'' vs.~``chemotherapy''). Then, the group terms in Equation \eqref{eq:linear-model} can be written as below with \(treatment_j=0\) representing the first treatment group (Group A) and \(treatment_j=1\) representing the second treatment group (Group B). With this notation, the linear model then can be expressed as

@@

\hypertarget{LMEM-case}{%
\subsubsection{The Linear Mixed Model (LMEM) Case}\label{LMEM-case}}

@@

Comparing the fitted responses of the LMEM and the rm-ANOVA models used in the simulated quadratic data (Figure \ref{fig:l-q-response}E, F) indicates that the models are not capturing the changes within each group. Specifically, note that the fitted mean response of both models shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured.

The LMEM is only able to account for between-subject variation by providing estimates for each subject (Figure \ref{fig:l-q-response}F), but both models are unable to capture the fact that the initial values are the same in each group, and instead fit non-parallel lines that have initial values that are markedly different from the ``true'' initial values in each case (compare Figure \ref{fig:l-q-response}D with Figure \ref{fig:l-q-response}E, F). If such a change has important physiological implications, both rm-ANOVA and LMEMs omit it from the fitted mean response. Thus, even though the model correctly detects a divergence between treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data. It could be argued that a LMEM with quadratic effects should have been used to fit the data in Figure\ref{fig:l-q-response}, F. However, because in reality the true function is not known, choosing a polynomial degree causes more questions (e.g., is it quadratic?, cubic?, or a higher degree?). Additionally, polynomial effects have other limitations, which we cover in Section \ref{GAM-theory}.

This section has used simulation to better convey the limitations of linearity and correlation in the response in non-linear data. The models fitted to the simulated data were an rm-ANOVA model and a LMEM, where the main issue is the expected linear trend in the response, which is only noticeable if the model fit and the response are visualized. In the following section, we provide a brief overview of linear models, general linear models and generalized linear mixed models before presenting the theory of GAMs.

\hypertarget{linear-models-and-beyond}{%
\section{Linear Models, and beyond}\label{linear-models-and-beyond}}

In reality, rm-ANOVA and LMEMs are just \emph{special cases} of broader model classes (General Linear Models and Generalized Linear Mixed Models, respectively). Understanding this is critical to fully capture the constraints of such models and to understand how GAMs overcome those limitations. This section will briefly present the different Statistical model classes, indicate how rm-ANOVA and LMEMs fit within that framework and introduce the theory of GAMs.

\hypertarget{linear-models-lms-and-linear-mixed-models-lmems}{%
\subsection{Linear Models (LMs) and Linear Mixed Models (LMEMs)}\label{linear-models-lms-and-linear-mixed-models-lmems}}

Linear models (LMs) are those that assume a normal (Gaussian) distribution of the response, such as rm-ANOVA. These are by far the models most commonly used to analyze data within the biomedical research community, and only incorporate \emph{fixed effects}. On the other hand, Linear Mixed Effect Models (LMEMs) are those that assume a normally-distributed response but also incorporate \emph{random effects}, as it has been described in Section \ref{LMEM-case}.

In matrix notation, a LM has the general form:

\begin{equation}
\textbf{y} = \textbf{X} \boldsymbol{\beta}
\label{eq:LM-matrix}
\end{equation}

Where \(\textbf{y}\) a vector of observations, \(\textbf{X}\) is a matrix that contains the measured values of the covariates, and \(\boldsymbol{\beta}\) is a vector of the parameters to be estimated. The ``model equation'' form of Equation \eqref{eq:LM-matrix} for a rm-ANOVA model of two treatment groups corresponds to Equation \eqref{eq:linear-model}. It is important to remember that the error term \(\varepsilon_{ijt}\) in Equation \eqref{eq:linear-model} is due to random sampling from the normal distribution.

\hypertarget{generalized-linear-models-glms}{%
\subsection{Generalized Linear Models (GLMs)}\label{generalized-linear-models-glms}}

A major limitation of LMs is their assumption of normality. If the data is non-normal, a transformation is necessary in order to properly fit the model. However, transformation lead to poor model performance,\textsuperscript{\protect\hyperlink{ref-ohara2010}{51}} and can cause problems with the biological interpretation of the model estimates. McCullagh and Nelder\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} defined General Linear Models (GLMs) as an extension of LMs, where the data does not need to be normally distributed. To achieve this, consider the following:

\begin{equation}
y_{ijt} \sim \mathcal{D}(\mu_{ijt},\phi)
\label{eq:GLM-y}
\end{equation}

Where \(y_{ijt}\) is the observation \(i\) in group \(j\) at time \(t\), which is assumed to come randomly from some distribution of the exponential family \(\mathcal{D}\), with some mean \(\mu_{ijt}\) and a dispersion parameter \(\phi\) (which in the Gaussian case is the variance \(\sigma^{2}\)). The mean \(\mu_{ijt}\) is also known as the \emph{expected value} (also called \emph{expectation}) (\(E\)) of the observed response. Then, the \emph{linear predictor} \(\eta\), which defines the relationship between the mean and the covariates can be defined as

\begin{equation}
\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}
\label{eq:GLM-eta}
\end{equation}

Where \(\eta_{ijt}\) is the linear predictor for each observation, \(\beta_0\) is the intercept, and \(\beta_{j}\) is the model parameter for each group; and \(x_{ijt}\) represents the covariates from each group at each time point.

Finally,

\begin{equation}
E(y_{ijt})=\mu_{ijt}=g^{-1}(\eta_{ijt})
\label{eq:GLM-Expectation}
\end{equation}
Where \(E(y_{ijt})\) is the expectation, \(g^-{1}\) is the inverse of the \emph{link function}, which relates the parameters of the distribution to the linear combination of the predictors in Equation \eqref{eq:GLM-eta}. Therefore, LMs (such as rm-ANOVA) are a special case of GLMs where the response is normally distributed.

\hypertarget{generalized-linear-mixed-models-glmms}{%
\subsection{Generalized linear mixed models (GLMMs)}\label{generalized-linear-mixed-models-glmms}}

Although GLMs relax the normality assumption, they can only accommodate fixed effects. Generalized Linear Mixed Models (GLMMs) are an extension of GLMs that incorporate \emph{random effects}, which have an associated probability distribution.\textsuperscript{\protect\hyperlink{ref-mcculloch2001}{53}} Therefore, in GLMMs the \emph{linear predictor} takes the form

\begin{equation}
\eta_{ijt}=\beta_0+\beta_{j}x_{ijt}+\mu_{ij}
\label{eq:GLMM-eta}
\end{equation}

Where \(\mu_{ij}\) corresponds to the random effects that can be estimated within each subject in each group, and all the other symbols correspond to the notation of the previous equations. Therefore, LMEMs are special case of GLMMs where the distribution of the response is normally distributed.\textsuperscript{\protect\hyperlink{ref-nelder1972}{52}} In-depth and excellent discussions about LMs, GLMs and GLMMs can be found in refs\textsuperscript{\protect\hyperlink{ref-dobson2008}{54}} and.\textsuperscript{\protect\hyperlink{ref-stroup2013}{55}}

\hypertarget{GAM-theory}{%
\subsubsection{GAMs as a special case of Generalized Linear Models}\label{GAM-theory}}

\hypertarget{gams-and-basis-functions}{%
\paragraph{GAMs and Basis Functions}\label{gams-and-basis-functions}}

Generalized additive models (GAMs) are a family of regression-based methods for estimating smoothly varying trends and are an extension of the GLM family.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-hastie1987}{56}} A GAM version of a linear model can be written as:

@@

Where \(y_{ijt}\) is the response at time \(t\) of subject \(i\) in group \(j\), \(\beta_0\) is the expected value at time 0, and where the change of \(y_{ijt}\) over time instead of being defined by \(\beta_{j}x_{ijt}\) as in Equation \eqref{eq:GLM-eta} is represented by the \emph{smooth function} \(f(x_t\mid \beta_j)\) with inputs as the covariates \(x_t\) and parameters \(\beta_j\), and \(\varepsilon_{ijt}\) represents the residual error.

@@

The smooth functional relationship between the covariates and the response in GAMs is specified using a semi-parametric relationship that can be fit within the GLM framework, by using \emph{basis function} expansions of the covariates and by estimating random coefficients associated with these basis functions. A \emph{basis} is a set of functions that spans the mathematical space within which the true but unknown \(f(x_t\mid \beta_j)\) is thought to exist.\textsuperscript{\protect\hyperlink{ref-simpson2018}{34}} For the linear model in Equation \eqref{eq:linear-model}, the basis coefficients are \(\beta_1\), \(\beta_2\) and \(\beta_3\) and the basis vectors are \(time_t\), \(treatment_j\) and \(time_t \times treatment_j\). The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response,\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}} which in the case of Equation \eqref{eq:linear-model} is restricted to a linear family of functions. In the case of Equation \eqref{eq:GAM}, the basis functions are contained in the expression \(f(x_t\mid \beta_j)\), which means that the model allows for non-linear relationships among the covariates.

Splines (which derive their name from the physical devices used by draughtsmen to draw smooth curves) are commonly used as \emph{basis functions}, as they have a long history in solving semi-parametric statistical problems and are often a default choice to fit GAMs as they are a simple, flexible and powerful option to obtain smoothness.\textsuperscript{\protect\hyperlink{ref-wegman1983}{58}} Although different types of splines exist, cubic, thin plate splines, and thin plate regression splines will be briefly discussed next to give a general idea of these type of basis functions, and their use within the GAM framework.

Cubic splines (CS), are smooth curves constructed from cubic polynomials joined together in a manner that enforces smoothness. The use of CS as smoothers in GAMs was discussed within the original GAM framework,\textsuperscript{\protect\hyperlink{ref-hastie1987}{56}} but they are limited by the fact that their implementation requires the selection of some points along the covariates (known as `knots,' the points where the bending of the smooth will occur) to obtain the reduced basis, which could affect the model fit.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} A solution to the ``knot'' placement of CS is provided by thin plate splines (TPS), which provide optimal smooth estimation without knot placement, but that are computationally costly to calculate.\textsuperscript{\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-wood2003}{59}}

In contrast, thin regression splines (TPRS) provide a reasonable ``low rank'' (truncated) approximation to the optimal TPS estimation, which can be implemented in an efficient computational manner.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} Like TPS, TPRS only require the number of basis to be used to create the smoother (for mathematical details on both TPS and TPRS see refs.\textsuperscript{\protect\hyperlink{ref-wood2003}{59}} and).\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}

@@

\hypertarget{a-bayesian-interpretation-of-gams}{%
\subsubsection{A Bayesian interpretation of GAMs}\label{a-bayesian-interpretation-of-gams}}

Bayes' theorem states that the probability of an event can be calculated using prior knowledge or belief.\textsuperscript{\protect\hyperlink{ref-mcelreath2018}{60}} In the case of data that shows non-linear trends, the belief that the \emph{true} trend of the data is likely to be smooth rather than extremely ``wiggly'' introduces the concept of a prior distribution for wiggliness (and therefore a Bayesian view) of GAMs.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} Moreover, GAMs are considered ``empirical'' Bayesian models when fitted using the package \emph{mgcv} because the smoothing parameters are estimated from the data (and not from a prior distribution as in the ``fully Bayesian'' case, which can be fitted using JAGS, Stan, or other probabilistic programming language).\textsuperscript{\protect\hyperlink{ref-miller2019}{61}} Therefore, the confidence intervals (CIs) calculated for the smooth terms using \emph{mgcv} are considered empirical Bayesian posterior credible intervals,\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33}} which have good ``frequentist'' coverage (point-wise coverage or ``single point'' coverage), and \emph{across the function} coverage.\textsuperscript{\protect\hyperlink{ref-wood2017}{37}}

@@

Contrary to this, an \emph{across the function} CI (like those estimated for GAMs using \emph{mgcv}) contains the true function through the entire range of the covariates (which would be time in the case of longitudinal data). For GAMs this means that if repeated samples are taken and a GAM and corresponding CIs are calculated, the percentage of CIs entirely containing the true function over the entire time period would be close to the nominal value (i.e., if 100 random samples are obtained and a GAM and CI is calculated for each one of them, it would be expected that 95 out of the 100 fitted CIs entirely contain the true function), thereby allowing more robust estimates from the model. In-depth theory of the Bayesian interpretation of GAMs is beyond the scope of this paper, but can be found in\textsuperscript{\protect\hyperlink{ref-simpson2018}{34},\protect\hyperlink{ref-wood2017}{37},\protect\hyperlink{ref-miller2019}{61}} and.\textsuperscript{\protect\hyperlink{ref-marra2012}{62}} With this brief introduction to the Bayesian interpretation of GAMs, we henceforth refer to the confidence intervals for the smooths in GAMs as ``empirical Bayesian'' through the rest of this paper.

@@

An interaction effect is typically the main interest in longitudinal biomedical data, as it takes into account treatment, time, and their combination. In a practical sense, when a GAM is implemented for longitudinal data, a smooth can be added to the model for the \emph{time} effect to account for the repeated measures over time. Although specific methods of how GAMs model correlation structures is a topic beyond the scope of this paper, it suffices to say that GAMs are flexible and can handle correlation structures beyond compound symmetry. A detailed description on basis functions and correlations can be found in.\textsuperscript{\protect\hyperlink{ref-hefley2017}{57}}

@@

When the smooths are plotted over the raw data, it is clear that the model has been able to capture the trend of the change of \(\mbox{StO}_2\) for each group across time (Figure \ref{fig:sim-smooth-plot}B). Model diagnostics can be obtained using the \passthrough{\lstinline!gam.check!} function, and the function \passthrough{\lstinline!appraise!} from the package \emph{gratia}.\textsuperscript{\protect\hyperlink{ref-gratia}{63}} A guide for model selection and diagnostics is in the \protect\hyperlink{workflow}{Appendix}, and an in-depth analysis can be found in\textsuperscript{\protect\hyperlink{ref-wood2017}{37}} and.\textsuperscript{\protect\hyperlink{ref-harezlak2018}{64}}

@@

This ``model misspecification'' error, also is known as a ``Type III'' error\textsuperscript{\protect\hyperlink{ref-dennis2019}{17}} is particularly important because although the \emph{p-value} is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model. We acknowledge that polynomial effects can be used in LMEMs, and that it could be argued that in Section \ref{simulation} a LMEM with quadratic effects could have been fitted, but in reality the true function in the data is not known; using polynomial effects causes more questions than answers in this case (i.e., what is the appropriate polynomial effect to use? quadratic?, cubic?, what is the biological interpretation of a polynomial effect?). Moreover, polynomial effects have other associated issues that we have described in Section \ref{GAM-theory}.

Guidelines for statistical reporting in biomedical journals exist (the SAMPL guidelines)\textsuperscript{\protect\hyperlink{ref-lang2015}{65}} but they have not been widely adopted and in the case of longitudinal data, we consider that researchers would benefit from reporting a visual assessment of the correspondence between the model fit and the data, rather than merely relying on a \(R^2\) value, which is not meaningful in the case of a Type III error.

In this paper we have presented GAMs as a suitable method to analyze longitudinal data with non-linear trends.\\
It is interesting to note that although GAMs are a well established method to analyze temporal data in different fields (among which are palaeoecology, geochemistry, and ecology)\textsuperscript{\protect\hyperlink{ref-pedersen2019}{33},\protect\hyperlink{ref-hefley2017}{57}} they are not routinely used in biomedical research despite an early publication from Hastie and Tibshirani that demonstrated their use in medical research.\textsuperscript{\protect\hyperlink{ref-hastie1995}{66}} This is possibly due to the fact that the theory behind GAMs can seem very different from that of rm-ANOVA and LMEMs, but the purpose of Section \ref{GAM-theory} is to demonstrate that at its core the principle is quite simple: Instead of using a linear relationship to model the response (as rm-ANOVA and LMEMs do), GAMs use basis functions to build smooths that are capable of learning non-linear trends in the data, which is a major advantage over models where the user has to know the non-linear relationship \emph{a priori} in order to provide it to the model (i.e., polynomial effects in LMEMs).

@@

We have used \(\textsf{R}\) as the software of choice for this paper because not only provides a fully developed environment to fit GAMs, but also eases simulation (which is becoming increasingly used for exploratory statistical analysis and power calculations) and provides powerful and convenient methods of visualization, which are key aspects that biomedical researchers might need to consider to make their work reproducible. In this regard, reproducibility is still an issue in biomedical research,\textsuperscript{\protect\hyperlink{ref-begley2015}{67},\protect\hyperlink{ref-weissgerber2018}{68}} but it is becoming apparent that what other disciplines have experienced in this aspect is likely to impact sooner rather than later this field. Researchers need to plan on how they will make their data, code, and any other materials open and accessible as more journals and funding agencies recognize the importance and benefits of open science in biomedical research. We have made all the data and code used in this paper accessible, and we hope that this will encourage other researchers to do the same with future projects.

@@

\leavevmode\vadjust pre{\hypertarget{ref-ohara2010}{}}%
\CSLLeftMargin{51. }
\CSLRightInline{OHara R, Kotze J. Do not log-transform count data. Published online January 2010. doi:\href{https://doi.org/10.1038/npre.2010.4136.1}{10.1038/npre.2010.4136.1}}

\leavevmode\vadjust pre{\hypertarget{ref-nelder1972}{}}%
\CSLLeftMargin{52. }
\CSLRightInline{Nelder JA, Wedderburn RWM. Generalized linear models. \emph{Journal of the Royal Statistical Society Series A (General)}. 1972;135(3):370. doi:\href{https://doi.org/10.2307/2344614}{10.2307/2344614}}

\leavevmode\vadjust pre{\hypertarget{ref-mcculloch2001}{}}%
\CSLLeftMargin{53. }
\CSLRightInline{McCulloch C. \emph{Generalized, Linear, and Mixed Models}. John Wiley \& Sons; 2001.}
